Bayesian exponential random graph modeling of KWL Burial network

### 2020-11-13 analysis/paper/paper.Rmd
edit background section

<!-- archaeological background -->
Archaeological sites in northeastern Taiwan show evidence of use of imported prestige goods in burials and some accumulation patterns in residential areas around the same time as European contact in the 17th century [@Chen2007]. This might indicate the pursuit of prestige or wealth in those Indigenous societies. Prestige goods include imported ornaments and porcelains that entered northeastern Taiwan via regional exchange networks that involved the Dutch, Spanish, Han Chinese traders, and local Indigenous groups. The use of prestige goods may have began in local Indigenous communities before a European presence became established, but was amplified during the European colonial period [@Borao2009; @LiandWu2006; @LWandBM2020ornament]. For example, an Indigenous person possessing more imported goods may have been recognized as more influential or having higher status by their community [@LiandWu2006]. A general model to summarize this situation in northeastern Taiwan might be that the influence of a colonial power, combined with high local values of imported goods, might lead to increased social inequality due to competition among individuals, and unequal access to trade networks [@Brumfiel1994; @Clark1994].

<!-- archaeological specific case study -->
In this study, we use a pericolonial model to explore the changes in social inequality at Kiwulan, an Iron Age site in northeastern Taiwan, which covers the time before the European arrival, the presence of the Europeans, including the Spanish and the Dutch, in the 17th century, and finally the Chinese in the 19th century [@Chen2007]. Our purpose is to examine the network structure of burials before and after the foreign contacts, and test our hypothesis that increased social inequality can be observed in the network after the European arrival. We ask: (1) does the observed burial data after the start of European presence resemble the of network our hypothesized model? (2) did European colonial activities in 17th century Taiwan result in the emergence of social inequality in an Indigenous society in ways that can be detected by burials networks? (3) what are the major variables affecting or forming the higher degree of popularity (e.g. a few nodes have more relationships with other nodes) in the European contact period that might hint at social heterogeneity during this period? By answering these questions, this study helps to expand the use of burials in understanding the indirect effects of a colonial presence on Indigenous groups.

# Exponential random graph models in a Baysesian framework

@@

<!--archaeological setting-->
We analyzed burial data collected from the upper component of Kiwulan, an Iron Age settlement (1350-1850 AD) located in northeastern Taiwan, which experienced European colonial impacts in the early 17th century and a large wave of Chinese immigrants in the 19th century [@Chen2007]. Excavations revealed abundant pottery sherds, imported ceramics and stonewares, wooden artifacts, stone tools, metal artifacts, imported glass beads and agates beads, and pipes [@Chen2007]. In addition to these artifacts, 90 burials, hundreds of middens, storage pits, and postholes with _in-situ_ posts were also excavated. The burials are mostly located in the middle section of the excavation area, which is the largest open area at Kiwulan that provides continuous stratigraphic sections suitable for temporal comparison <!-- fig reference to a map of burials at the site-->. Burials are oriented in an east-west direction on the north side of the residential area, indicated by post-holes and `in-situ` wooden posts, indicating a well organized spatial arrangement. Previous studies report an uneven distribution of prestige goods across burials without agreement about whether this uneven distribution hints at vertical social differences. For example, @Cheng2008 interpreted the unequal distribution of glass beads, especially the gold-foil beads between burials, as evidence for hierarchy, indicating a stratified society. However, @Hsieh2012 suggested a relatively egalitarian structure based on comparative analysis of the frequencies of all burial goods. She found that the burials with high value burial goods were usually associated with elders, which might indicate achieved, rather than inherited, status. One important limitation of these previous studies is that they did not use analytical units suitable for comparing pre-European social organiation with post-European social organisation. Here, we adopt a new chronological framework for the burials to test if network configurations differ from the pre-European period to the post-European period.

<!-- take the citations from this paragraph and add them to one of the intro paras, them delete this para -->
The earliest record of the use of prestige goods in burials can be traced back to the time when the European arrival in the 17th century according to historical records from the Spanish [@Borao2009; @LiandWu2006]. For example, the document written by a Spanish Priest described that the Spanish soldier used carnelian beads to exchange natural resources with local Indigenous people since their high values in Indigenous culture [@LiandWu2006]. In addition, there was a custom that people would bury imported beads and ceramics with deceased to display their status and influence [@LiandWu2006]. Despite we observed the uneven distribution of prestige goods in burials, the degree of differentiation across burials over time remain unclear. The discord over the discussion of Kiwulan social organization could be associated with chronological differences that has not yet been well studied. That is, the observed uneven distribution may be explained as a result of social change from less social inequality to more social inequality when Indigenous societies in Northeastern Taiwan were involved in the complex trade network stimulated by the Europeans. We assume that the changes would be supported by chronological differences in the structure of burial networks. Analyzing and comparing networks between different phases relative to cultural contact will give an insight into changes in social differences. This is important to understand the reactions of Kiwulan residents in the context of indirect impacts of European colonization.

@@

<!-- we should tell the reader what does it mean high-medium-low --> 
We analyzed burial data collected from the published excavation reports, and the original fieldwork notes of Kiwulan. Gold-foil beads and agate beads are commonly found across burials with substantial differences in quantities, so we described each burial as having one of three levels, high (50-1000?), medium (20-50?), and low (1-10?), of these beads. That means if burial 1 and burial 2 both have high quantities of agate beads, then there will be a tie connecting them. For less frequent prestige goods, including imported porcelains, gold foils, and fish-shaped ornaments, we linked two burials when they possess each type of goods in common (i.e. presence or absence). Node attributes here include osteological data, such as age, sex, and ritual pottery, and an overall burial goods values... index. Ritual pottery was identified as locally made ceramics placed above graves that suggests funeral feasting according to historical records. The quantity index is an important attribute for economical inequality that we assigned the burials into three classes, high, medium, and low, as an index of wealth, based on the total number of burial goods. Since burials tend to have multiple prestige goods in common, the network ties are weighted instead of binary data (where the value 1 represents a tie and the value 0 otherwise) [@Snijders2011]. For example, if two burials have both low quantity of glass beads and porcelain in common, the tie is given a value of 2. Our networks are non-directed, which means ties have no orientation forming the relationship between actors to indicate a mutual relationship.

>>>>

<!-- archaeological background -->
Archaeological sites in northeastern Taiwan show evidence of imported prestige goods, such as ornaments and porcelains, in burials and some accumulation patterns in residential areas around the same time as European contact in the 17th century [@Chen2007; @LWandBM2020ornament]. The use of imported goods is also mentioned in historical records from the Spanish, indicative of the pursuit of prestige or wealth in those Indigenous societies [@Borao2009; @LiandWu2006]. For example, the document written by a Spanish Priest described that the Spanish soldier used carnelian beads to exchange natural resources with local Indigenous people since their high values in Indigenous culture [@LiandWu2006]. In addition, an Indigenous person possessing more imported goods may have been recognized as more influential or having higher status by their community [@LiandWu2006]. Despite we observed the uneven distribution of prestige goods in burials, the degree of differentiation across burials over time remain unclear. The introduction of prestige goods to local Indigenous communities might occur earlier before a European presence became established, but was amplified during the European colonial period [@Borao2009; @LiandWu2006; @LWandBM2020ornament]. A general model to summarize this situation in northeastern Taiwan might be that the influence of a colonial power, combined with high local values of imported goods, might lead to increased social inequality due to competition among individuals, and unequal access to trade networks [@Brumfiel1994; @Clark1994]. The observed uneven distribution may be explained as a result of social change from less social inequality to more social inequality when Indigenous societies in Northeastern Taiwan were involved in the complex trade network stimulated by the Europeans.

<!-- archaeological specific case study -->
In this study, we use a pericolonial model to explore the changes in social inequality at Kiwulan, an Iron Age site in northeastern Taiwan, which covers the time before the European arrival, the presence of the Europeans, including the Spanish and the Dutch, in the 17th century, and finally the Chinese in the 19th century [@Chen2007]. We assume that the social changes at Kiwulan would be supported by chronological differences in the structure of burial networks. We examine burial networks before and after the foreign contacts, and test our hypothesis that increased social inequality can be observed in the network after the European arrival. We ask: (1) does the observed burial data after the start of European presence resemble the of network our hypothesized model? (2) did European colonial activities in 17th century Taiwan result in the emergence of social inequality in an Indigenous society in ways that can be detected by burials networks? (3) what are the major variables affecting or forming the higher degree of popularity (e.g. a few nodes have more relationships with other nodes) in the European contact period that might hint at social heterogeneity during this period? By answering these questions, this study helps to expand the use of burials in understanding the indirect effects of a colonial presence on Indigenous groups. This is important to understand the reactions of Kiwulan residents in the context of indirect impacts of European colonization. 
# Exponential random graph models in a Baysesian framework

@@

<!--archaeological setting-->
We analyzed burial data collected from the upper component of Kiwulan, an Iron Age settlement (1350-1850 AD) located in northeastern Taiwan, which experienced European colonial impacts in the early 17th century and a large wave of Chinese immigrants in the 19th century [@Chen2007]. Excavations revealed abundant pottery sherds, imported ceramics and stonewares, wooden artifacts, stone tools, metal artifacts, imported glass beads and agates beads, and pipes [@Chen2007]. In addition to these artifacts, 90 burials, hundreds of middens, storage pits, and postholes with _in-situ_ posts were also excavated. The burials are mostly located in the middle section of the excavation area, which is the largest open area at Kiwulan that provides continuous stratigraphic sections suitable for temporal comparison <!-- fig reference to a map of burials at the site-->. Burials are oriented in an east-west direction on the north side of the residential area, indicated by post-holes and `in-situ` wooden posts, indicating a well organized spatial arrangement. Previous studies report an uneven distribution of prestige goods across burials without agreement about whether this uneven distribution hints at vertical social differences. For example, @Cheng2008 interpreted the unequal distribution of glass beads, especially the gold-foil beads between burials, as evidence for hierarchy, indicating a stratified society. However, @Hsieh2012 suggested a relatively egalitarian structure based on comparative analysis of the frequencies of all burial goods. She found that the burials with high value burial goods were usually associated with elders, which might indicate achieved, rather than inherited, status. One important limitation of these previous studies is that they did not use analytical units suitable for comparing pre-European social organization with post-European social organization. Here, we adopt a new chronological framework for the burials to test if network configurations differ from the pre-European period to the post-European period. The discord over the discussion of Kiwulan social organization could be associated with chronological differences that has not yet been well studied.

@@

<!-- we should tell the reader what does it mean high-medium-low --> 
We analyzed burial data collected from the published excavation reports, and the original fieldwork notes of Kiwulan. Gold-foil and agate beads are commonly found across burials with substantial differences in quantities, so we described each burial as having one of three levels, high (> 10), medium (> 1 and < 10), and low (= 1), of these beads. Glass beads are in levels of high (> 100), medium (> 99 and < 900), and low (< 100). That means if burial 1 and burial 2 both have high quantities of agate beads, then there will be a tie connecting them. For less frequent prestige goods, including imported porcelains, gold foils, and fish-shaped ornaments, we linked two burials when they possess each type of goods in common (i.e. presence or absence). Node attributes here include osteological data, such as age and sex, and cultural data, such as ritual pottery, and burial value index. Ritual pottery was identified as locally made ceramics placed above graves that suggests funeral feasting according to historical records. The burial value index is an important attribute for economical inequality that we assigned the burials into three classes, high (>30), medium(>11 and <30), and low(<12), as an index of wealth. The number is the sum of values of each type of prestige good, which is calculate by the total number of a prestige item from burial context over the number of prestige item in a burial. Since burials tend to have multiple prestige goods in common, the network ties are weighted instead of binary data (where the value 1 represents a tie and the value 0 otherwise) [@Snijders2011]. For example, if two burials have both low quantity of glass beads and porcelain in common, the tie is given a value of 2. Our networks are non-directed, which means ties have no orientation forming the relationship between actors to indicate a mutual relationship.
